Psychological Needs During Intergroup Contact: Three Experience Sampling Studies

### 2022-03-15 Methods-and-Results.tex
M & R feedback

After receiving ethical approval from the University of Groningen, we
recruited 23 migrants using the local paid participant pool and
specifically targeted non-Dutch migrants to participate in our study.
Participants reported on their interactions for at least 30 days with
two daily measures (capturing the morning and afternoon). With this
design, we aimed at getting 50-60 measurements per participant
(\textit{M} = 53.26, \textit{SD} = 16.72, \textit{total N} = 1225). This
is a common number of measurements found in experience sampling studies
and offers sufficient power to model processes within and between
participants \citep[e.g., for a systematic review see][]{AanhetRot2012}.
Participants were compensated for their participation with up to 34
Euros -- each two Euros for pre- and post-questionnaire as well as 50
Eurocents for every experience sampling measurement occasion. The sample
consisted of relatively young, educated, and western migrants from the
global north (\(M_{age}\) = 24.35, \(SD_{age}\) = 4.73, 19 women, 15
students). The sample accurately describes the largest groups of
migrants in the region \citep[][]{GemeenteGroningen2015}.

@@

The study itself consisted of three main parts, an introductory
pre-measurement, and the daily experience sampling measurements, as well
as a concluding post-measurement. After giving informed consent,
participants started by filling in an online pre-questionnaire assessing
demographics and general information about their immigration. Over the
next thirty days, the participants were then invited twice a day (at 12
pm and 7pm) to reflect upon their interactions, psychological need
fulfillments, and current attitudes towards the Dutch outgroup
(\textit{median duration} = 2M 22S, \textit{MAD duration} = 1M 19S).
General compliance was high (85.90\% of all invited surveys were filled
in)\footnote{Two participants completed only two days (among the others, participation was 93.70\%)}.
The response rates were approximately equal during mornings (\textit{n}
= 621) and afternoons (\textit{n} = 604) and most measurements were
completed within four hours of the invitation. After the final day of
daily diary measurements, participants were invited to fill in a longer
post measurement survey that mirrored the pre-measurement. All key
variables for this study were part of the short daily diary surveys.

@@

To test the prerequisite effect of intergroup contact, every experience
sampling measurement started with the question
``\textit{Did you meet a Dutch person this morning [/afternoon]? (In person interaction for at least 10 minutes)}''.
Our participants recorded between 2--51 interactions with Dutch outgroup
members (3.23--91.07\% of individual daily diary measurements; 31.59\%
of all 1225 daily diary
responses)\footnote{Two participants only recorded two daily diary measurements each and non of these included outgroup contacts. These participants are removed from any analyses that focus on outgroup contacts.}.

@@

At the end of every daily diary measurement we asked all participants
about their current attitudes towards the Dutch -- our main dependent
variable. To assess the momentary outgroup evaluation we used the common
feeling thermometer: ``How favorable do you feel towards the Dutch?''
\citep[][]{Lavrakas2008}. Participants then rated their attitude on a
continuous slider scale from ``very cold -- 0'' through ``no feeling --
50'' to ``very warm -- 100''. Both the question phrasing as well as the
tick labels were consistent with large-scale panel surveys
\citep[e.g.,][]{DeBell2010}.

@@

We tested the most general contact hypothesis in two steps. First, we
assessed whether more intergroup interactions were related to to more
positive outgroup attitudes. Second, we tested whether a potential
positive effect on outgroup attitudes depended on the interaction
quality (jointly with the number of interactions). We find that neither
the number of interactions nor the number of daily diary responses with
an interaction were significantly related with the average outgroup
attitudes (all \textbar{} \textit{r} \textbar{} \textless{} 0.03, all
\textit{p} \textgreater{} 0.887). However, we did find a medium sized
correlation between the participants' Average Interaction Quality and
their Average Outgroup Attitudes (\textit{r} = 0.55, \textit{p} = 0.010.
And when considering the number of interactions and average interaction
quality jointly in a linear regression, we additionally find a
statistically significant interaction term (\textit{b} = -0.05,
\textit{t}(17) = -2.95, \textit{p} = 0.009, \(\eta_p^2\) = 0.34), where
in our sample with an increasing number of interactions the positive
effect of average interaction quality becomes weaker (also see Table
\ref{tab:intergroupGeneralTblLong}). We additionally used a multilevel
regression to check whether having an interaction with an outgroup
member had a situational (i.e., contemporaneous) effect within the
participants. We find that having an outgroup interaction is indeed
associated with significantly more positive outgroup attitudes within
the participants (\textit{b} = 2.48, \textit{t}(1200) = 4.36, \textit{p}
\textless{} .001, \textit{95\%CI}{[}1.36, 3.59{]}), even after
controlling for having an interaction with a non-Dutch (which did not
relate to outgroup attitudes independently; For full results see Table
\ref{tab:intergroupGeneralTblLong} and Online Supplementary Materials
B)\footnote{Interestingly, adding random slopes to this model did not explain additional variance. This is unusual and might indicate the the effect is very consistent across participants. However, the small number of participants, or other measurement issues provide an alternative explanation}.
Thus, in our first data we find mixed results, where outgroup contacts
show a positive effect on outgroup attitudes in the moment but these
results do not translate into average differences between participants
(even tough the average number of interactions does seem to interact
with average interaction quality to predict outgroup attitudes).
However, it should be noted that the between-participant analyses in the
aggregated form are based on a small sample size.

@@

The main proposal of our article is that the success of an outgroup
interaction might be explained by whether or not the interaction
fulfilled the person's core situational need. This should, in turn, be
due to a higher perceived interaction quality. We sequentially test
whether the fulfillment of the core need during an interaction is (1)
related to more positive outgroup attitudes, (2) higher perceived
interaction quality, and (3) whether the variance explained by the core
need is assumed by the perceived interaction quality if considered
jointly (for full regression results see Table
\ref{tab:intergroupNeedsTblLong}).

We find that in the highest multilevel models, the fulfillment of core
situational needs during outgroup contacts was associated with more
positive outgroup attitudes (random slopes model; \textit{b} = 0.17,
\textit{t}(365) = 2.93, \textit{p} = 0.004, \textit{95\%CI}{[}0.06,
0.29{]}) and also predicted higher perceived interaction quality (random
intercept model; \textit{b} = 0.42, \textit{t}(365) = 8.07, \textit{p}
\textless{} .001, \textit{95\%CI}{[}0.32, 0.52{]}). Moreover, when we
consider the influences of core need fulfillment and interaction quality
on outgroup attitudes jointly, we find that virtually all variance is
explained by perceived interaction quality (random slopes mode;
\textit{b} = 0.23, \textit{t}(364) = 4.22, \textit{p} \textless{} .001,
\textit{95\%CI}{[}0.12, 0.33{]}) and only little unique variance is
still explained by core need fulfillment (\textit{b} = 0.04,
\textit{t}(364) = 0.64, \textit{p} = 0.523, \textit{95\%CI}{[}-0.08,
0.15{]}). We thus find support for our hypotheses and can conclude that
in this data set the fulfillment of core situational needs had a
significant influence on outgroup attitudes. Additionally, this effect
is likely being explained by its effect through perceived interaction
quality.

@@

To build further confidence in our results, we assessed two additional
models that might offer alternative explanations. First, to ensure that
the effect of core need fulfillment is specific to an actual
interaction, we compared the effect to core need fulfillement in
situations without an interaction. For this, we analyzed the generalized
situational core need fulfillment (either during an interaction or about
the daytime in general) and tested whether the effect differed during
daily diary measurements with and without outgroup contacts. We found no
main effect of core need fulfillment (random slopes model; \textit{b} =
-0.10, \textit{t}(1199) = -2.22, \textit{p} = 0.026,
\textit{95\%CI}{[}-0.18, -0.01{]}) but a significant interaction effect
of core need fulfillment and outgroup contact (\textit{b} = 0.13,
\textit{t}(1199) = 4.31, \textit{p} \textless{} .001,
\textit{95\%CI}{[}0.07, 0.18{]}; also see Table
\ref{tab:robustnessTblLong}). Together with a significant main effect of
having an outgroup contact, this indicates that it is not key need
fulfillment in general --- but only key need fulfillment during an
outgroup contact that predicts more positive outgroup attitudes.

In a final step, we controlled for other fundamental psychological needs
during the interaction. We focus on the three commonly considered self
determination needs: competence, autonomy, and relatedness. We find that
the core need adds significantly above a model with only the self
determination theory needs (random slopes models; \(\chi^2\)(6,
\textit{N} = 21) = 19.60, \textit{p} = 0.003). We find that next to
relatedness (\textit{b} = 0.10, \textit{t}(361) = 4.04, \textit{p}
\textless{} .001, \textit{95\%CI}{[}0.05, 0.14{]}), the core need
explains the most variance in outgroup attitudes after an outgroup
contact (\textit{b} = 0.09, \textit{t}(361) = 2.23, \textit{p} = 0.026,
\textit{95\%CI}{[}0.01, 0.18{]}). When compared to the model with only
the SDT needs, the core need fulfillment flexibly takes on some of the
explained variance of all three fundamental needs (competence and
autonomy needs turning non-significant; all \textit{b} \textless{} 0.05,
all \textit{p} \textgreater{} 0.544). For full results see see Table
\ref{tab:robustnessTblLong} and Online Supplementary Information B.

@@

After receiving ethical approval from the University of Groningen, we
recruited 113 international students using a local participant pool. We
specifically targeted non-Dutch students, who had recently arrived in
the Netherlands. Participants reported on their interactions for at
least 30 days with two daily measures (capturing the morning and
afternoon). With this design, we aimed at getting 50-60 measurements per
participant (\textit{M} = 43.94, \textit{SD} = 15.00, \textit{total N} =
4965). As with the previous study this should offer sufficient power to
model processes within participants and will lend stronger weight to
between-participant results. Participants were compensated for their
participation with partial course credits --- depending on their
participation. The sample consisted of relatively young migrants, who
were mostly from the global north (\(M_{age}\) = 20.24, \(SD_{age}\) =
2.12, 84 women). The sample fairly accurately describes the local
population of international students.

@@

The study procedure mirrored the setup of study one and consisted of
pre-, daily diary-, and post-measurement. The participants were invited
for daily diary measurements twice a day (at 12 pm and 7pm) for 30 days
(\textit{median duration} = 3M 46S, \textit{MAD duration} = 1M 54S).
General compliance was high (70.87\% of all invited surveys were filled
in). The response rates were approximately equal during mornings
(\textit{n} = 2608) and afternoons (\textit{n} = 2357). All key
variables for this study were part of the short daily diary surveys.

@@

To measure intergroup contacts, every experience sampling measurement
started with the question
``\textit{Did you meet a Dutch person this morning [/afternoon]? (in-person interaction for at least 10 minutes)}''.
Participants were additionally offered a pop-up explanation: ``With
in-person interaction, we mean a continued interaction with another
person (potentially in a group) that lasted at least 10 minutes. This
interaction should be offline and face-to-face. It should include some
form of verbal communication and should be uninterrupted to still count
as the same interaction. Any individual interaction can last minutes or
hours. If there were multiple interaction partners, we would like you to
focus on the person that was most important to you during the
interaction.''. The participants recorded between 1--43 interactions
with Dutch majority people (1.64--93.75\% of individual daily diary
measurements; 18.83\% of all 4965 daily diary responses).

@@

For the core situational need, we asked participants in an open ended
text field:
``\textit{What was your main goal [during the interaction with X / this morning / this afternoon]?}''.
Participants could additionally click on a pop-up explanation: ``Your
main goal during an interaction can vary depending on the interaction.
It could be to connect with friends, to find or provide help, to achieve
academic ambitions, work on your fitness, work for a job, or simply to
get a coffee, just as well as many many other concrete or abstract goals
that are import to you in the moment. It really depends on your
subjective experience of the interaction.''. Then, with reference to the
text entry, we asked how much this core need was fulfilled during the
interaction or the past daytime period:
``\textit{During your interaction with X [this morning / this evening] your goal (-previous text entry-) was fulfilled.}''
on a continuous slider scale ranging from strongly disagree (1) to
strongly agree (100).

The measurement of the self determination needs was identical to Study
1.

@@

The ratings of the perceived interaction quality was identical to study
one.

@@

As in Study 1, attitudes towards the Dutch majority outgroup were again
measured using the feeling thermometer. All survey details are also
available in Online Supplemental Materials A and B.

@@

We tested the most general contact hypothesis in the same manner as in
Study 1. We found a significant correlation with average outgroup
attitudes for both the total number of outgroup interactions (\textit{r}
= 0.22, \textit{p} = 0.019) and the number of measurement beeps with an
interaction (\textit{r} = 0.28, \textit{p} = 0.003). However, we found
no significant correlation between the participants' Average Interaction
Quality and their Average Outgroup Attitudes (\textit{r} = 0.04,
\textit{p} = 0.679). Thus, when considering the number of interactions
and average interaction quality jointly in a linear regression, we find
a main effect of intergroup contacts to predict outgroup attitudes
(\textit{b} = 0.81, \textit{t}(109) = 3.29, \textit{p} = 0.001,
\(\eta_p^2\) = 0.08) but no significant effect of average perceived
contact quality and no significant interaction term (for full results
see Table \ref{tab:intergroupGeneralTblLong}). We additionally used a
multilevel regression to check whether having an interaction with an
outgroup member had a situational (i.e., contemporaneous) effect within
the participants. We find that having an outgroup interaction is indeed
associated with significantly more positive outgroup attitudes within
the participants (random slopes model; \textit{b} = 2.99,
\textit{t}(4850) = 3.80, \textit{p} \textless{} .001,
\textit{95\%CI}{[}1.45, 4.54{]}), even after controlling for having an
interaction with a non-Dutch person (which did not relate to outgroup
attitudes independently; For full results see Table
\ref{tab:intergroupGeneralTblLong} and Online Supplementary Materials
B). Thus, in our second data set we find mixed results, outgroup
contacts show a positive effect on outgroup attitudes in the moment and
on average between participants. However, we find that this effect does
not depend on the average perceived interaction quality. This unexpected
results, might be due to the aggregation process and the following
analyses will focus in on the role of the perceived interaction quality
during an outgroup interaction.

@@

We again sequentially tested whether the fulfillment of the core need
during an interaction was (1) related to more positive outgroup
attitudes, (2) higher perceived interaction quality, and (3) whether the
variance explained by the core need is assumed by the perceived
interaction quality if considered jointly. We find that in the
multilevel models, the fulfillment of core situational needs during
outgroup contacts was associated with more positive outgroup attitudes
(random slopes model; \textit{b} = 0.13, \textit{t}(826) = 4.18,
\textit{p} \textless{} .001, \textit{95\%CI}{[}0.07, 0.19{]}) and also
predicted higher perceived interaction quality (random slopes model;
\textit{b} = 0.40, \textit{t}(826) = 6.46, \textit{p} \textless{} .001,
\textit{95\%CI}{[}0.28, 0.52{]}). Additionally, if we consider the
influences of core need fulfillment and interaction quality on outgroup
attitudes jointly, we find that virtually all variance is explained by
perceived interaction quality (random slopes mode; \textit{b} = 0.17,
\textit{t}(825) = 5.56, \textit{p} \textless{} .001,
\textit{95\%CI}{[}0.11, 0.23{]}) and only little unique variance is
still explained by core need fulfillment (\textit{b} = 0.03,
\textit{t}(825) = 1.47, \textit{p} = 0.141, \textit{95\%CI}{[}-0.01,
0.07{]}; also see Table \ref{tab:intergroupNeedsTblLong} for full
results). These results are consistent with the results in study one and
we thus find support for our hypotheses and can conclude that in this
data set the fulfillment of core situational needs had a significant
influence on outgroup attitudes and that this effect is likely explained
by its effect through perceived interaction quality.

@@

We again checked for alternative models. First, when considering
generalized situational core need fulfillment together with whether an
intergroup contact took place, we find that there is only a minuscule
main effect of core need fulfillment (random slopes model; \textit{b} =
-0.03, \textit{t}(4849) = -1.34, \textit{p} = 0.181,
\textit{95\%CI}{[}-0.08, 0.02{]}) but a stronger interaction effect of
core need fulfillment and outgroup contact (\textit{b} = 0.06,
\textit{t}(4849) = 3.03, \textit{p} = 0.002, \textit{95\%CI}{[}0.02,
0.10{]}). Together with a significant main effect of having an outgroup
contact (\textit{b} = 2.88, \textit{t}(4849) = 3.71, \textit{p}
\textless{} .001, \textit{95\%CI}{[}1.36, 4.40{]}), this indicates that
it is not key need fulfillment in general --- but key need fulfillment
during an outgroup contact that predicts more positive outgroup
attitudes. This finding is consistent with the results of the previous
study, albeit with slightly weaker effect (likely because of the large
number of measurements that did not include an outgroup interaction; For
full results see Table \ref{tab:intergroupNeedsTblLong} and for visual
comparison see Figure \ref{fig:AllportNeedFulfillment}).

In a final step we again checked whether during the interaction the core
situational need remains a meaningful predictor even when taking other
fundamental psychological needs into account. We find that the core need
adds significantly above a model with only the self determination theory
needs (random slopes models; \(\chi^2\)(6, \textit{N} = 108) = 22.90,
\textit{p} \textless{} .001). We find that the core need explains the
most variance in outgroup attitudes after an outgroup contact
(\textit{b} = 0.08, \textit{t}(823) = 2.95, \textit{p} = 0.003,
\textit{95\%CI}{[}0.03, 0.13{]}). When compared to the model with only
the SDT needs, the core need fulfillment flexibly takes on some of the
explained variance of all of the three fundamental needs. However,
different from the first study, relatedness (\textit{b} = 0.07,
\textit{t}(823) = 4.26, \textit{p} \textless{} .001,
\textit{95\%CI}{[}0.04, 0.10{]}) and autonomy (\textit{b} = 0.02,
\textit{t}(823) = 0.93, \textit{p} = 0.353, \textit{95\%CI}{[}-0.02,
0.08{]}) also predicted positive outgroup attitudes in this larger
sample. For full results see Table \ref{tab:intergroupNeedsTblLong} and
Online Supplementary Information B.

@@

The ratings of the perceived interaction quality was identical to study
one.

>>>>

After receiving ethical approval from the University Masked for Peer
Review, we recruited 23 migrants using the local paid participant pool
and specifically targeted non-Dutch migrants to participate in our
study. Participants reported on their interactions for at least 30 days
with two daily measures (capturing the morning and afternoon). With this
design, we aimed at getting 50-60 measurements per participant
(\textit{M} = 53.26, \textit{SD} = 16.72, \textit{total N} = 1,225).
This is a common number of measurements found in experience sampling
studies and offers sufficient power to model processes within and
between participants
\citep[e.g., for a systematic review see][]{AanhetRot2012}. Participants
were compensated for their participation with up to 34 Euros -- each two
Euros for pre- and post-questionnaire as well as 50 Eurocents for every
experience sampling measurement occasion. The sample consisted of
relatively young, educated, and western migrants from the global north
(\(M_{age}\) = 24.35, \(SD_{age}\) = 4.73, 19 women, 15 students). The
sample accurately describes the largest groups of migrants in the region
\citep[][]{GemeenteGroningen2015}.

@@

The study itself consisted of three main parts, an introductory
pre-measurement, and the daily experience sampling measurements, as well
as a concluding post-measurement. After giving informed consent,
participants started by filling in an online pre-questionnaire assessing
demographics and general information about their immigration. Over the
next thirty days, the participants were then invited twice a day (at 12
pm and 7pm) to reflect upon their interactions, psychological need
fulfillments, and current attitudes towards the Dutch outgroup
(\textit{median duration} = 2M 22S, \textit{MAD duration} = 1M 19S).
General compliance was high (85.90\% of all invited surveys were filled
in)\footnote{Two participants completed only two days (among the others, participation was 93.70\%).}.
The response rates were approximately equal during mornings (\textit{n}
= 621) and afternoons (\textit{n} = 604) and most measurements were
completed within four hours of the invitation. After the final day of
experience sampling measurements, participants were invited to fill in a
longer post measurement survey that mirrored the pre-measurement. All
key variables for this study were part of the short experience sampling
surveys.

@@

To test the prerequisite effect of intergroup contact, every experience
sampling measurement started with the question
``\textit{Did you meet a Dutch person this morning [/afternoon]? (In person interaction for at least 10 minutes)}''.
Our participants recorded between 2--51 interactions with Dutch outgroup
members (3.23--91.07\% of individual experience sampling measurements;
31.59\% of all 1,225 experience sampling
responses)\footnote{Two participants only recorded two experience sampling measurements each and non of these included outgroup contacts. These participants are removed from any analyses that focus on outgroup contacts.}.

@@

At the end of every experience sampling measurement we asked all
participants about their current attitudes towards the Dutch -- our main
dependent variable. To assess the momentary outgroup evaluation we used
the common feeling thermometer: ``How favorable do you feel towards the
Dutch?'' \citep[][]{Lavrakas2008}. Participants then rated their
attitude on a continuous slider scale from ``very cold -- 0'' through
``no feeling -- 50'' to ``very warm -- 100''. Both the question phrasing
as well as the tick labels were consistent with large-scale panel
surveys \citep[e.g.,][]{DeBell2010}.

@@

We tested the most general contact hypothesis in two steps. First, we
assessed whether overall between participants more intergroup
interactions were related to to more positive average outgroup
attitudes. Second, we tested whether a potential positive effect on
outgroup attitudes depended on the interaction quality (jointly with the
number of interactions). We find that neither the overall number of
contacts nor the number of experience sampling responses with a contact
were significantly related with the average outgroup attitudes (all
\textbar{} \textit{r} \textbar{} \textless{} 0.03, all \textit{p}
\textgreater{} 0.887). However, we did find a medium sized correlation
between the participants' average contact quality and their average
outgroup attitudes (\textit{r} = 0.55, \textit{p} = 0.010). And when
considering the number of contacts and average contact quality jointly
in a linear regression, we additionally found a statistically
significant interaction term (\textit{b} = -0.05, \textit{t}(17) =
-2.95, \textit{p} = 0.009, \(\eta_p^2\) = 0.34), where in our sample
with an increasing number of contacts the positive effect of average
contact quality becomes weaker (also see Table
\ref{tab:intergroupGeneralTblLong}). We additionally used a multilevel
regression to check whether having an interaction with an outgroup
member had a situational (i.e., contemporaneous) effect within the
participants. We find that having an outgroup contact is indeed
associated with significantly more positive outgroup attitudes within
the participants (\textit{b} = 2.48, \textit{t}(1,200) = 4.36,
\textit{p} \textless{} .001, \textit{95\%CI}{[}1.36, 3.59{]}), even
after controlling for having an interaction with a non-Dutch (which did
not relate to outgroup attitudes independently; For full results see
Table \ref{tab:intergroupGeneralTblLong}, Figure
\ref{fig:ContactHypothesis}, and Online Supplementary Material
A)\footnote{Interestingly, adding random slopes to this model did not explain additional variance. This is unusual and might indicate that the effect is very consistent across participants. However, the small number of participants, or other measurement issues provide an alternative explanation.}.
Thus, in our first data we find mixed results, where outgroup contacts
show a positive effect on outgroup attitudes in the moment but these
results do not translate into average differences between participants
(even tough the average number of contacts does seem to interact with
average contact quality to predict outgroup attitudes). However, it
should be noted that the between-participant analyses in the aggregated
form are based on a small sample size.

@@

The main proposal of our article is that the success of an outgroup
contact might be explained by whether or not the contact fulfilled the
person's core situational need. This should, in turn, be due to a higher
perceived contact quality. We sequentially test whether the fulfillment
of the core need during an interaction is (1) related to more positive
outgroup attitudes, (2) higher perceived contact quality, and (3)
whether the variance explained by the core need is subsumed by the
perceived contact quality if considered jointly (for full regression
results see Table \ref{tab:intergroupNeedsTblLong} and Figure
\ref{fig:AllportNeedFulfillment}).

We find that in the multilevel models, the fulfillment of core
situational needs during outgroup contacts was associated with more
positive outgroup attitudes (random slopes model; \textit{b} = 0.17,
\textit{t}(365) = 2.93, \textit{p} = 0.004, \textit{95\%CI}{[}0.06,
0.29{]}) and also predicted higher perceived contact quality (random
intercept model; \textit{b} = 0.42, \textit{t}(365) = 8.07, \textit{p}
\textless{} .001, \textit{95\%CI}{[}0.32, 0.52{]}). Moreover, when we
consider the influences of core need fulfillment and contact quality on
outgroup attitudes jointly, we find that virtually all variance is
explained by perceived contact quality (random slopes mode; \textit{b} =
0.23, \textit{t}(364) = 4.22, \textit{p} \textless{} .001,
\textit{95\%CI}{[}0.12, 0.33{]}) and only little unique variance is
still explained by core need fulfillment (\textit{b} = 0.04,
\textit{t}(364) = 0.64, \textit{p} = 0.523, \textit{95\%CI}{[}-0.08,
0.15{]}). We thus find support for our hypotheses and can conclude that
in this data set the fulfillment of core situational needs had a
significant influence on outgroup attitudes. Additionally, this effect
is likely explained by its effect through perceived contact quality.

@@

To build further confidence in our results, we assessed two additional
models that might offer alternative explanations. First, to ensure that
the effect of core need fulfillment is specific to an actual contact, we
compared the effect to core need fulfillment in situations without an
intergroup contact. For this, we analyzed the generalized situational
core need fulfillment (either during a contact or about the daytime in
general) and tested whether the effect differed during experience
sampling measurements with and without outgroup contacts. We found no
main effect of core need fulfillment (random slopes model; \textit{b} =
-0.10, \textit{t}(1,199) = -2.22, \textit{p} = 0.026,
\textit{95\%CI}{[}-0.18, -0.01{]}) but a significant interaction effect
of core need fulfillment and outgroup contact (\textit{b} = 0.13,
\textit{t}(1,199) = 4.31, \textit{p} \textless{} .001,
\textit{95\%CI}{[}0.07, 0.18{]}; also see Table
\ref{tab:robustnessTblLong} and Figure \ref{fig:Robustness}). Together
with a significant main effect of having an outgroup contact, this
indicates that it is not key need fulfillment in general --- but only
key need fulfillment during an outgroup contact that predicts more
positive outgroup attitudes.

In a final step, we controlled for other fundamental psychological needs
during the contact. We focus on the three commonly considered
self-determination needs (SDT): competence, autonomy, and relatedness.
We find that the core need fulfillment adds significantly above a model
with only the self-determination theory needs (random slopes models;
\(\chi^2\)(6, \textit{N} = 21) = 19.60, \textit{p} = 0.003). We also
find that next to relatedness (\textit{b} = 0.10, \textit{t}(361) =
4.04, \textit{p} \textless{} .001, \textit{95\%CI}{[}0.05, 0.14{]}), the
core need explains the most variance in outgroup attitudes after an
outgroup contact (\textit{b} = 0.09, \textit{t}(361) = 2.23, \textit{p}
= 0.026, \textit{95\%CI}{[}0.01, 0.18{]}). When compared to the model
with only the SDT needs, the core need fulfillment flexibly takes on
some of the explained variance of all three fundamental needs
(competence and autonomy needs turning non-significant; all \textit{b}
\textless{} 0.05, all \textit{p} \textgreater{} 0.544). For full results
see see Table \ref{tab:robustnessTblLong}, Figure \ref{fig:Robustness},
and Online Supplementary Material A.

@@

We recruited 113 international students using a local participant pool.
We specifically targeted non-Dutch students, who had recently arrived in
the Netherlands. Participants reported on their interactions for at
least 30 days with two daily measures (capturing the morning and
afternoon). With this design, we again aimed at receiving 50-60
measurements per participant (\textit{M} = 43.94, \textit{SD} = 15.00,
\textit{total N} = 4,965). As with the previous study this should offer
sufficient power to model processes within participants and will lend
stronger weight to between-participant results. Participants were
compensated for their participation with partial course credits ---
depending on their participation. The sample consisted of relatively
young migrants, who were mostly from the global north (\(M_{age}\) =
20.24, \(SD_{age}\) = 2.12, 84 women). The sample fairly accurately
describes the local population of international students.

@@

The study procedure mirrored the setup of Study 1 and consisted of pre-,
experience sampling-, and post-measurements. The participants were
invited for experience sampling measurements twice a day (at 12 pm and
7pm) for 30 days (\textit{median duration} = 3M 46S,
\textit{MAD duration} = 1M 54S). General compliance was high (70.87\% of
all invited surveys were filled in). The response rates were
approximately equal during mornings (\textit{n} = 2,608) and afternoons
(\textit{n} = 2,357). All key variables for this study were part of the
short experience sampling surveys.

@@

To measure intergroup contacts, every experience sampling measurement
started with the question
``\textit{Did you meet a Dutch person this morning [/afternoon]? (in-person interaction for at least 10 minutes)}''.
Participants were additionally offered a pop-up explanation: ``With
in-person interaction, we mean a continued interaction with another
person (potentially in a group) that lasted at least 10 minutes. This
interaction should be offline and face-to-face. It should include some
form of verbal communication and should be uninterrupted to still count
as the same interaction. Any individual interaction can last minutes or
hours. If there were multiple interaction partners, we would like you to
focus on the person that was most important to you during the
interaction.''. The participants recorded between 1--43 interactions
with Dutch majority people (1.64--93.75\% of individual experience
sampling measurements; 18.83\% of all 4,965 experience sampling
responses).

@@

For the core situational need, we asked participants in an open ended
text field:
``\textit{What was your main goal [during the interaction with -X- / this morning / this afternoon]?}''
(where \textit{-X-} was dynamically replaced with the name of the
interaction partner). Participants could additionally click on a pop-up
explanation: ``Your main goal during an interaction can vary depending
on the interaction. It could be to connect with friends, to find or
provide help, to achieve academic ambitions, work on your fitness, work
for a job, or simply to get a coffee, just as well as many other
concrete or abstract goals that are import to you in the moment. It
really depends on your subjective experience of the interaction.''.
Then, with reference to the text entry, we asked how much this core need
was fulfilled during the interaction or the past daytime period:
``\textit{During your interaction with -X- [this morning / this evening] your goal (-previous text entry-) was fulfilled.}''
on a continuous slider scale ranging from strongly disagree (1) to
strongly agree (100).

The measurement of the self-determination needs was identical to Study
1.

@@

The ratings of the perceived contact quality was identical to Study 1.

@@

As in Study 1, attitudes towards the Dutch majority outgroup were again
measured using the feeling thermometer. All survey details are also
available in our OSF repository \citep{KreienkampMasked2022a} and Online
Supplemental Materials A.

@@

We tested the most general contact hypothesis as in Study 1. We found a
significant correlation with average outgroup attitudes for both the
total number of outgroup interactions (\textit{r} = 0.22, \textit{p} =
0.019) and the number of measurement beeps with an interaction
(\textit{r} = 0.28, \textit{p} = 0.003). However, we found no
significant correlation between the participants' average interaction
quality and their average outgroup attitudes (\textit{r} = 0.04,
\textit{p} = 0.679). Thus, when considering the number of interactions
and average interaction quality jointly in a linear regression, we find
a main effect of intergroup contacts to predict outgroup attitudes
(\textit{b} = 0.81, \textit{t}(109) = 3.29, \textit{p} = 0.001,
\(\eta_p^2\) = 0.08) but no significant effect of average perceived
contact quality and no significant interaction term (for full results
see Table \ref{tab:intergroupGeneralTblLong} and Figure
\ref{fig:ContactHypothesis}). We additionally used a multilevel
regression to check whether having an interaction with an outgroup
member had a situational (i.e., contemporaneous) effect within the
participants. We find that having an outgroup interaction is indeed
associated with significantly more positive outgroup attitudes within
the participants (random slopes model; \textit{b} = 2.99,
\textit{t}(4,850) = 3.80, \textit{p} \textless{} .001,
\textit{95\%CI}{[}1.45, 4.54{]}), even after controlling for having an
interaction with a non-Dutch person (which did not relate to outgroup
attitudes independently; For full results see Table
\ref{tab:intergroupGeneralTblLong}, Figure \ref{fig:ContactHypothesis},
as well as Online Supplementary Material A). Thus, in our second data
set we find mixed results; outgroup contacts show a positive effect on
outgroup attitudes in the moment and on average between participants.
However, we find that this effect does not depend on the average
perceived interaction quality. This unexpected results, might be due to
the aggregation process and the following analyses will focus in on the
role of the perceived interaction quality during an outgroup
interaction.

@@

We again sequentially tested whether the fulfillment of the core need
during an interaction was (1) related to more positive outgroup
attitudes, (2) higher perceived interaction quality, and (3) whether the
variance explained by the core need is subsumed by the perceived
interaction quality if considered jointly. We find that in the
multilevel models, the fulfillment of core situational needs during
outgroup contacts was associated with more positive outgroup attitudes
(random slopes model; \textit{b} = 0.13, \textit{t}(826) = 4.18,
\textit{p} \textless{} .001, \textit{95\%CI}{[}0.07, 0.19{]}) and also
predicted higher perceived interaction quality (random slopes model;
\textit{b} = 0.40, \textit{t}(826) = 6.46, \textit{p} \textless{} .001,
\textit{95\%CI}{[}0.28, 0.52{]}). Additionally, if we consider the
influences of core need fulfillment and interaction quality on outgroup
attitudes jointly, we find that virtually all variance is explained by
perceived interaction quality (random slopes mode; \textit{b} = 0.17,
\textit{t}(825) = 5.56, \textit{p} \textless{} .001,
\textit{95\%CI}{[}0.11, 0.23{]}) and only little unique variance is
still explained by core need fulfillment (\textit{b} = 0.03,
\textit{t}(825) = 1.47, \textit{p} = 0.141, \textit{95\%CI}{[}-0.01,
0.07{]}; also see Table \ref{tab:intergroupNeedsTblLong} and Figure
\ref{fig:AllportNeedFulfillment} for full results). These results are
consistent with the results in Study 1. We, thus, find support for our
hypotheses that the fulfillment of core situational needs had a
significant influence on outgroup attitudes and that this effect is
likely explained by its effect through perceived interaction quality.

@@

We again checked for alternative models. First, when considering
generalized situational core need fulfillment together with whether an
intergroup contact took place, we find that there is only a minuscule
main effect of core need fulfillment (random slopes model; \textit{b} =
-0.03, \textit{t}(4,849) = -1.34, \textit{p} = 0.181,
\textit{95\%CI}{[}-0.08, 0.02{]}) but a stronger interaction effect of
core need fulfillment and outgroup contact (\textit{b} = 0.06,
\textit{t}(4,849) = 3.03, \textit{p} = 0.002, \textit{95\%CI}{[}0.02,
0.10{]}). Together with a significant main effect of having an outgroup
contact (\textit{b} = 2.88, \textit{t}(4,849) = 3.71, \textit{p}
\textless{} .001, \textit{95\%CI}{[}1.36, 4.40{]}), this indicates that
it is not key need fulfillment in general --- but key need fulfillment
during an outgroup contact that predicts more positive outgroup
attitudes. This finding is consistent with the results of the previous
study, albeit with a slightly weaker effect (likely because of the large
number of measurements that did not include an outgroup interaction; For
full results see Table \ref{tab:intergroupNeedsTblLong} and for visual
comparison see Figure \ref{fig:AllportNeedFulfillment}).

In a final step we again checked whether during the interaction the core
situational need remains a meaningful predictor even when taking other
fundamental psychological needs into account. We find that the core need
fulfillment adds significantly above a model with only the
self-determination theory needs (random slopes models; \(\chi^2\)(6,
\textit{N} = 108) = 22.90, \textit{p} \textless{} .001). We find that
the core need fulfillment explained the most variance in outgroup
attitudes after an outgroup contact (\textit{b} = 0.08, \textit{t}(823)
= 2.95, \textit{p} = 0.003, \textit{95\%CI}{[}0.03, 0.13{]}). When
compared to the model with only the SDT needs, the core need fulfillment
flexibly takes on some of the explained variance of all of the three
fundamental needs. However, different from the first study, relatedness
(\textit{b} = 0.07, \textit{t}(823) = 4.26, \textit{p} \textless{} .001,
\textit{95\%CI}{[}0.04, 0.10{]}) and autonomy (\textit{b} = 0.02,
\textit{t}(823) = 0.93, \textit{p} = 0.353, \textit{95\%CI}{[}-0.02,
0.08{]}) also predicted positive outgroup attitudes in this larger
sample. For full results see Table \ref{tab:intergroupNeedsTblLong},
Figure \ref{fig:Robustness}, and Online Supplementary Material A.

@@

The ratings of the perceived interaction quality was identical to Study
1.
